Models

### 2021-04-08 analysis/05_model_metrics.Rmd

metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1),
         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values

```
####Data exploration

How many years are missing FD metrics? Only a handful! Let's exclude them so we can get an accurate view of the number of years for which we have data for each timeseries                                                                                                     
```{r}
no_fd_years <- metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))

dim(no_fd_years)

metrics <- setdiff(metrics, no_fd_years)
```

@@

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  mutate(study_id = substr(rarefyID, 1, 3))

@@

SES_rand_slope <- model_data %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)))#,
  #          null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
  #                                data = .x)),
  #          test = map2(model, null_model, anova),
  #          test_res = map(test, tidy),
  #          coef = map(model, tidy)
  #          ) %>% 
  # unnest(cols = c(test_res)) %>%
  # select(metric, model, coef, p.value) %>%
  # filter(!is.na(p.value)) %>%
  # unnest(cols = c(coef)) %>%
  # filter(term == "year_scaled") %>%
  # select(-c(effect, group, term, statistic))

@@

>>>>

#get metadata
meta <- pins::pin_get("meta", board = "github") %>%
  select(study_id, realm, climate, habitat, protected_area, biome_map)

meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  select(study_id, rarefyid, taxa) %>% distinct() %>%
  left_join(meta)

#get metric data
metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  left_join(meta, c("rarefyID" = "rarefyid"))
```

####Data exploration

How many years are missing FD metrics? Only a handful! Let's exclude them so we can get an accurate view of the number of years for which we have data for each timeseries                                                                                                     
```{r}
# no_fd_years <- metrics %>%
#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))
# 
# dim(no_fd_years)
# 
# metrics <- setdiff(metrics, no_fd_years)
```

@@

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`)

@@

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>%
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

Clustering 
```{r}
scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)

blup_mat <- blup %>%  
  #these studies have NA's for some reason, remove them 
  filter(!study_id %in% c(445, 446, 348, 382)) %>%
  select(study_id, starts_with("SES"), S) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

clmethods <- c("hierarchical","kmeans","pam")
intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")

summary(intern)
optimalScores(stab)

#get the hierarchical clustering object
clust <- intern@clusterObjs$hierarchical
cat5 <- cutree(clust, k = 5)
cat3 <- cutree(clust, k = 3)

blup_clust <- blup %>% 
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```
